% !TEX root = ./thesis.tex
\graphicspath{{./figures/vega-arch/}}
\chapter{A Streaming Dataflow Architecture}
\label{sec:vg:arch}

While the previous chapter describes the design of Reactive Vega's declarative
interaction design model, we now turn to the system architecture needed to
support it. Our architecture design is motivated by four primary goals:

\begin{enumerate}
  \item \textbf{A Unified Data Model}. Existing reactive visualization
toolkits (e.g., Model.js~\cite{kelleher:modeljs}) feature fragmented
architectures where only interaction events are modeled as time-varying. Other
input datasets remain static and batch-processed. This artificial disconnect
restricts expressivity and can result in wasteful computation. For example,
interaction events that manipulate only a subset of input tuples may trigger
recomputation over the entire dataset. In contrast, Reactive Vega features a
unified data model in which input data, scene graph elements, and interaction
events are all treated as first-class streaming data sources.

  \item \textbf{Streaming Relational Data}. Modeling input relational data with
Event-Driven Functional Reactive Programming (E-FRP)~\cite{wan:efrp} semantics
alone does not supply sufficient granularity for targeted recomputation. As
E-FRP semantics consider only time-varying scalar values, operators would
observe an entire relation as having changed and so would need to reprocess all
tuples. Instead, Reactive Vega integrates techniques from streaming
databases~\cite{abadi:borealis, abadi:aurora, arasu:stream, avnur:eddies,
chandrasekaran:telegraphcq} alongside E-FRP, including tracking state at the
tuple-level and only propagating modified tuples through the dataflow graph.

  \item \textbf{Streaming Nested Data}. Interactive visualizations, particularly
those involving small multiples, often require hierarchical structures.
Processing such data poses an additional challenge not faced by prior reactive
or streaming database systems. To support streaming nested data, Reactive Vega's
dataflow graph rewrites itself in a data-driven fashion at runtime: new branches
are extended, or existing branches pruned, in response to observed hierarchies.
Each dataflow branch models its corresponding part of the hierarchy as a
standard relation, enabling operators to remain agnostic to higher-level
structure.

  \item \textbf{Interactive Performance}. Reactive Vega performs both compile-
and run-time optimizations to increase throughput and reduce memory footprint,
including tracking metadata to prune unnecessary computation, and optimizing
scheduling by inlining linear chains of operators. We conduct benchmark
studies of streaming and interactive visualizations and find that Reactive
Vega meets or exceeds the performance of both D3 and the original, unreactive
Vega system.
\end{enumerate}

Reactive Vega is implemented in the JavaScript programming language, and is
intended to run either in a web browser or server-side using Node.js. By
default, Reactive Vega renders to an HTML5 Canvas element; however, it also
supports Scalable Vector Graphics (SVG) and server-side image rendering.

\input{vega-arch/dataflow}
\input{vega-arch/optimizations}
\input{vega-arch/performance}
\input{vega-arch/conclusion}