% !TEX root = ./thesis.tex
\chapter{Related Work}
\label{sec:related_work}

This dissertation builds on prior work in visualization systems, programming
languages, streaming databases, and human-computer interaction. In this chapter,
I survey pertinent work in each domain and describe how this thesis extends
them.

\section{Visualization Toolkits \& Systems}

Chart templates, as found in spreadsheets and online services (e.g., Many
Eyes~\cite{ibm:manyeyes} and Google Fusion Tables), are a common means for
creating visualizations: users select from a pre-defined palette of chart types.
Such \emph{chart typologies} are popular because they are simple to use and
efficient\,---\,users can produce a recognizable chart with only a few clicks.
This approach, however, is closed. Users are restricted to only the available
chart types, and may customize only a small handful of parameters.

To enable custom visualization design, researchers have developed programming
toolkits that offer a number of visualization components. For instance, the
InfoVis Toolkit~\cite{fekete:ivtk} provides a class hierarchy of visualization
widgets, where new visualizations are created by subclassing existing components
or writing new ones. In contrast, Prefuse~\cite{heer:prefuse} and
Flare~\cite{flare} offer composable operators for data transform, layout, and
encoding. While either approach offers fine-grained control over visual output,
they both constrain expressivity much like chart typologies. Novel designs
require new widgets or operators, and significant software engineering effort.

\subsection{Grammar-Based Visual Encoding}

In 1999, Leland Wilkinson introduced an alternative approach with \emph{The
Grammar of Graphics}~\cite{wilkinson:grammar}. Eschewing chart typologies in
favor of a combinatorial building blocks, Wilkinson's grammar yields a more
expressive design space and allows users to rapidly construct visualizations.
Provided abstractions include primitives for data modeling and transformation,
visual encodings defined as mappings between data fields and channels (e.g.,
position, color, and size), and scales and guides (i.e., axes and legends). His
work was quickly followed by the Stanford Polaris system~\cite{stolte:polaris}
(commercialized as Tableau), and Hadley Wickham's popular
ggplot2~\cite{wickham:layered} package implements a variant of this model in the
R statistical language.

These tools focus on statistical graphics for exploratory analysis and, to that
end, favor concise specification languages. Analysts may omit details such as
scale transforms (e.g., linear or log) or color palettes, which are then filled
in by the grammar using a rule-based system of smart defaults. As a result,
analysts are able to rapidly generate visualizations, analyze them, and continue
with their analysis process.

Through the lower-level Protovis~\cite{bostock:protovis,heer:protovisjava} and
D3~\cite{bostock:d3} languages, Bostock and Heer describe how a similar approach
can enable expressive design for custom, explanatory graphics as well. Perhaps
most importantly, they articulate the advantages of declarative design for data
visualizations: by separating specification from execution, declarative
languages facilitate an iterative development process, can transparently
optimize processing, and simplify retargeting across
platforms~\cite{heer:protovisjava}.

The design of Reactive Vega and Vega-Lite is heavily influenced by these works.
Reactive Vega's visual encoding primitives and implementation pipeline closely
mirror Protovis'. Marks including \emph{arcs}, \emph{areas}, \emph{bars},
\emph{lines}, plotting \emph{symbols} and \emph{text} are processed through
\emph{bind}, \emph{build}, and \emph{evaluate} stages to generate a scene graph.
Protovis' \emph{panel} marks are renamed to \emph{group} marks, and enable the
repeated and nested structures found in small multiple displays. Marks can be
positioned relative to one-another with Vega's \emph{reactive geometry}, which
provides a more unified approach to layout as compared with Protovis'
\emph{anchors}.

Similarly, Vega-Lite represents basic plots using a set of encoding definitions
that map data attributes to visual channels, and may include common data
transformations. Vega-Lite differs most from other high-level grammars in its
approach to multiple view displays. Each of these grammars supports faceting (or
nesting) to construct trellis plots in which each cell similarly visualizes a
different partition of the data. Both Wilkinson's grammar and Polaris/Tableau
achieve this through a \emph{table algebra} over data fields, which in turn
determines spatial subdivisions. Tableau additionally supports the construction
of multi-view dashboards via a different mechanism, with each view backed by a
separate specification. In contrast, Vega-Lite contributes a \emph{view
algebra}: starting with unit specifications that define a single plot, Vega-Lite
expresses composite views using operators for layering, horizontal or vertical
concatenation, faceting, and parameterized repetition. When applicable, these
operators will merge scale domains and properly align constituent views.
Disparate views can also be combined into arbitrary dashboards, all within a
unified algebraic model.

\subsection{Dataflow Visualization Systems}

Dataflow architectures are common in scientific visualization systems, such as
IBM Data Explorer~\cite{lucas:dataflow,abram:ibmdx} and
VTK~\cite{schroeder:vtk}. Developers must manually specify and connect each
required operator into a network, which can support updates in a demand-driven
fashion (e.g., as data is modified) or an event-driven fashion (e.g., in
response to user input). These systems expose fine-grained control over the
construction of the dataflow graph. For example, VTK developers can choose to
favor memory efficiency over processing speed, which causes dataflow operators
to delete their output after computation. While Reactive Vega shares some
dataflow strategies with these systems\,---\,for example, using
pass-by-reference for unchanged tuples to reduce memory consumption\,---\,it
abstracts such execution concerns away from the user. The dataflow graph is
automatically assembled based on definitions found in a declarative
specification, and optimizations are transparently performed such that output
data is only stored when needed by downstream operators and shared otherwise.

Within the domain of information visualization, the Stencil
language~\cite{cottam:stencil} is also grounded in reactive semantics and uses a
dataflow model. Like Reactive Vega, it provides a unified data model where both
input data and interaction events are modeled as first-class streaming data
sources. However, Reactive Vega is more expressive than Stencil in two important
ways. First, Stencil does not provide primitives, beyond event streams, to
support interaction design; Reactive Vega, on the other hand, offers predicates
and scale inversions to lift interactive selections into data queries, and
coordinate multiple displays. Moreover, graphical primitives can be arbitrarily
nested with Reactive Vega, drawing from either hierarchical or distinct data
sources. This ability is critical to concisely specifying small multiples
displays, and requires Reactive Vega's dataflow graph to dynamically rewrite
itself at runtime. To the best of our knowledge, Stencil's architecture does not
support self-instantiating dataflows.

Improvise~\cite{weaver:improvise} features active variables called ``live
properties,'' which provide a basic form of reactivity\,---\,they may be bound
to control widgets and parameterize a visualization. Using an expression
language, live properties are assembled into a coordination graph to dynamically
evaluate visual encodings and generate views of data. While Improvise and
Reactive Vega share some conceptual underpinnings, Improvise places a higher
burden on users to correctly construct the necessary graph. As Reactive Vega
takes a declarative approach to visualization design, users need only compose
the necessary primitives into a specification. Reactive Vega parses this
specification to build the corresponding dataflow graph.

\subsection{Interactive Visualization Systems}

Besides text-based toolkits, researchers have also developed graphical user
interfaces for constructing visualizations.
Polaris/Tableau~\cite{stolte:polaris}, for instance, displays a ``schema'' panel
that lists available data fields. Data groupings and visual encodings are
specified by dragging these fields onto ``shelves'' that array the periphery of
the visualization. Lyra extends this mode of interaction onto the visualization
canvas itself\,---\,property \emph{drop zones} overlay marks during drag
operations, providing a more direct data-mapping target.

Roth et al.'s SageBrush~\cite{roth:sagebrush} supports similar interactions.
Users can drag-and-drop ``partial prototypes'' for spatial encodings and
``grapheme'' (mark) primitives such as lines and bars. Custom grapheme
properties are manually selected via menus, then exposed as drop-target icons.
Lyra refines the SageBrush model in several ways. Partial prototypes in
SageBrush implicitly define scale transforms and layout; by leveraging the Vega
and Vega-Lite grammars, Lyra decouples these components to support richer
designs. Lyra's data pipelines provide an extensible set of data
transformations, and different marks can be driven by unique pipelines and
composed. Finally, Lyra eschews iconic abstractions in favor of direct drop
zones for mark properties, without need of explicit enumeration via menus.

More recently, Bret Victor demonstrated a data-driven drawing
tool~\cite{victor:drawing} (now developed as
Apparatus\footnote{http://aprt.us/}) that combines geometric constraints with
\emph{imperative} procedures over data. Alongside an interactive canvas and data
table viewer, the tool includes programming structures with lexical scope and
control flow via loops and conditionals. To form a basic bar chart, designers
must define loops to create and position each bar. The tool supports purely
geometric constructions, rendering some layouts inexpressible. By building on
Vega and Vega-Lite, Lyra takes a \emph{declarative} approach to design. When a
designer associates a mark with a dataset, one mark instance per datum is
automatically produced rather than requiring explicitly instantiation by the
designer. More advanced algorithms, which are difficult to express with Victor's
tool, are accessible as part of Lyra's graphical data pipeline.

\section{Specifying Interactions in Visualizations}

Despite the central role of interaction in effective data visualization
\cite{heer:dynamics, pike:interactionscience}, little work has been done to
develop a grammar for specifying interaction techniques. Wilkinson's grammar
includes no notion of interaction. Tableau supports common interaction
techniques, but relies on mechanisms external to the visual encoding grammar.
Early systems like GGobi~\cite{swayne:ggobi} support common techniques as well,
and provide imperative APIs for custom methods. However, such APIs make easy
tasks needlessly complex, burdening developers with learning low-level execution
details. More recent systems, including Protovis, D3, and
VisDock~\cite{choi:visdock}, offer a typology of common techniques that can be
applied to a visualization. Such top-down approaches, however, limit
customization and composition. For example, D3's interactors encapsulate event
processing, making it difficult to combine them if their events conflict (e.g.,
if dragging triggers brushing \emph{and} panning).

\subsection{Interactive Selection \& Querying}

Selection, often in the form of users clicking or lassoing visual items, is a
fundamental operation in user interfaces and has been well-studied in the
context of data visualization. For example, in Snap-Together
Visualization~\cite{north:snap}, multiple views are coordinated via ``primary-''
and ``foreign-key actions,'' which propagate selected data tuples from one view
to the others. Wilhelm~\cite{wilhelm:interaction} describes the need for such
``indirect object manipulation'' methods as an axiom of interactive data
displays. Chen's compound brushing~\cite{chen:compound} provides a visual
dataflow language for specifying a rich space of transformations of brush
selections.  More recently, Brunel~\cite{brunel} provides a special
\texttt{\#selection} data field that is dynamically populated with the elements
a user interacts with, and can be used to link multiple views or filter input
data. Similarly, RStudio's Shiny~\cite{shiny}, an imperative web application
layer, provides \texttt{brushedPoints} and \texttt{nearestPoints} functions
which can be used to operate on selected elements.

Other systems have studied formally representing selections as data
queries~\cite{wilhelm:interaction}. For example, brushing interactions in
VQE~\cite{derthick:interactive} generate \emph{extensional} queries that
enumerate all items of interest; a form-based interface enables specification of
\emph{intensional} (declarative) queries. Individual point and brush selections
in DEVise~\cite{livny:devise}, known as \emph{visual queries}, map to a
declarative structure and are used to link together multiple views. With
VIQING~\cite{olsten:viqing}, rectangular ``rubber band'' selections are modeled
as range extents, and views can be dropped on top of each other to join their
underlying datasets. Heer et al.~\cite{heer:generalized} demonstrate that by
modeling a selection as a declarative query, interactive ``query relaxation''
can successively capture more items of interest.

Reactive Vega and Vega-Lite both build on this work with abstractions for
constructing selections. In Reactive Vega, \emph{predicates} express interactive
selections in visual space, are lifted to data queries by inverting them through
scale functions, and participate in if-then-else chains to manipulate visual
encoding rules. Vega-Lite abstracts these primitives with a higher-level
\emph{selection}. Vega-Lite selections are populated with one or more points of
interest, in response to user interaction. Their built-in predicates map to
declarative queries, thereby allowing a minimal set of ``backing'' points to
represent the full space of selected points. Additional operators can transform
a selection's predicate or backing points (e.g., offsetting them to translate a
brush selection or perform panning). Selections parameterize visual encodings by
driving conditional logic, serving as input data, or defining scale extents.

\section{Alternate Paradigms for Imperative Callbacks}

For custom interaction design with existing visualization toolkits, users must
author imperative event handling callbacks which present several development
hurdles~\cite{myers:callbacks}. Callbacks registered outside the visualization
specification allow users to externally manipulate visualization state and
manually update dependencies~\cite{cooper:embedding}. This approach breaks
encapsulation and necessitates the introduction of side-effects: external calls
to maintain state~\cite{cooper:integrating}. Callback execution order can also
be unpredictable, requiring users to coordinate interleaved
calls~\cite{edwards:coherent}. As a result, callbacks can make it difficult to
reason about the current state of the visualization, which is now the product of
both the original specification and the side-effects of interaction callbacks.

\subsection{Constraint Programming}

One alternative to callback-oriented imperative programming is constraint-based
specification. Systems like Gilt~\cite{myers:callbacks} and
$\mu$constraints~\cite{hudson:constraints} implement one-way constraints: when
the value of a constrained interface is modified, its dependents are
automatically affected. Other systems, like Cassowary~\cite{badros:cassowary},
implement two-way constraints using more complex constraint solvers. Recently,
ConstraintJS~\cite{oney:constraintjs} and Bret Victor's prototype for drawing
data visualizations~\cite{victor:drawing} have shown that constraint programming
is suitable for authoring web applications and data visualizations,
respectively.

However, while constraint systems remove many of the obstacles introduced by
callbacks, they do not involve general consideration of event handling\,---\,a
crucial element for interaction design. In fact, the authors of ConstraintJS
intend for their system to complement event
architectures~\cite{oney:constraintjs}.

\subsection{Functional Reactive Programming}

Functional Reactive Programming (FRP) formalizes semantics similar to one-way
constraints. Drawing from dataflow programming, FRP recasts mutable values as
continuous, time-varying data streams~\cite{bainomugisha:frpsurvey}. We focus on
a discrete variant called Event-Driven FRP (E-FRP)~\cite{wan:efrp}. To capture
value changes as they occur, E-FRP provides \emph{streams}, which are infinite
time-ordered sequences of discrete events. Streams can be composed into
\emph{signals} to build expressions that react to events. The E-FRP runtime
constructs the necessary dataflow graph such that, when a new event fires, it
propagates to corresponding streams. Dependent signals are evaluated in a
two\--phase update: in the first phase, signals use prior computed values of
their dependents, which are subsequently updated in the second phase. E-FRP has
been shown to be viable for authoring interactive web
applications~\cite{czaplicki:elm, meyerovich:flapjax} and
visualizations~\cite{cottam:stencil, kelleher:modeljs}.

Reactive Vega's declarative interaction primitives remain grounded in E-FRP
semantics, and they preserve the two-phase update: interdependent signals are
updated in the order in which they are defined in the specification. However,
naive applications of E-FRP to visualization tasks can result in wasteful
recomputation. Traditional E-FRP primitives support only \emph{scalar} values,
whereas visualization pipelines must also process relational and hierarchical
data. Modeling these latter data types as scalar values provides insufficient
granularity to perform targeted recomputation. To efficiently support relational
data, Reactive Vega integrates streaming database methods; and, to support
streaming hierarchical data, the dataflow graph dynamically rewrites itself at
runtime, instantiating new branches to process nested relations.

\section{Data Stream Management}

The problem of managing streaming data has been well studied in the database
community. Researchers have developed an arsenal of techniques through the
development of systems such as Aurora~\cite{abadi:aurora},
Eddies~\cite{avnur:eddies}, STREAM~\cite{arasu:stream}, and
TelegraphCQ~\cite{chandrasekaran:telegraphcq}. As tuples are observed by these
systems, they are flagged as either new or removed. Tuples, rather than full
relations, are passed between operators in a query plan (realized as a dataflow
graph). As a result, operators can inspect just the updated tuples to perform
efficient computation. However, for some operations a set of changed tuples is
insufficient. For example, a join of two relations requires access to all tuples
within a specified window. In such cases, caches (sometimes referred to as
\emph{views}~\cite{abadi:aurora} or \emph{synopses}~\cite{arasu:stream}) are
used to materialize a relation, and shared among dependent operators.

Borealis~\cite{abadi:borealis} extends this work in two ways. To support
streaming tuple modifications, it introduces a revision processing scheme. An
operator can be \emph{replayed} with revised tuples in place of the original
data, and will then only emit corresponding revisions. Similarly, to enable
dynamic operator parameters, Borealis introduces \emph{time travel}. When an
operator parameter changes, an \texttt{undo} is issued to the nearest cache. The
cache emits tuple deletions, effectively ``rewinding'' the system to a previous
time. A subsequent \texttt{replay} triggers recomputation with the new parameter
value.

However, existing streaming data systems concern flat relations. Reactive Vega
instantiates these techniques, alongside E-FRP, within a visualization pipeline
and extends them to support streaming nested data. To do so, Reactive Vega's
dataflow graph dynamically rewrites itself at runtime with new branches. These
branches unpack nested relations, enabling downstream operators to remain
agnostic to higher-level structure while supporting arbitrary levels of nesting.

\section{Generative Models of User Interfaces}

Michel Beaudoin-Lafon's instrumental interaction
paradigm~\cite{beaudouin:instrumental} helped motivate this dissertation's
approach of considering declarative interaction design at varying levels of
abstraction. He identifies three properties of interaction models for post-WIMP
interaction: \emph{descriptive} of existing and new techniques;
\emph{comparative} to evaluate alternative designs; and, \emph{generative}, to
facilitate the identification and creation of new points in the design space.
Reactive Vega only offers the first property\,---\,an expressive design
space\,---\,but is too low-level to be generative or facilitate evaluation of
similar behaviors. Vega-Lite, on the other hand, offers both an expressive and
generative design space, suitable for exploring alternative techniques as
demonstrated in~\Cref{sec:vl:examples}. Kim et al. also demonstrate its
comparative powers, using Vega-Lite to model optimal sequences of
visualizations~\cite{kim:graphscape}.

Moreover, our articulation of Vega-Lite as a parametric design language is
inspired by Mackinlay's APT~\cite{mackinlay:apt} and his follow-on work with
Card and Robertson conducting a morphological analysis of input
devices~\cite{card:morphological}. In particular, we sought to identify and
decouple orthogonal axes of the design space. While \Cref{sec:vl:examples}
begins to systematically consider points along these axes, a full morphological
analysis of the Vega-Lite design space is left to future work.