% !TEX root = ./thesis.tex
\graphicspath{{./vega-lang/figures/}}
\chapter{Declarative Primitives for Interaction Design}
\label{sec:vg:lang}

Reactive Vega builds on a long-running thread of research on declarative
visualization design, popularized by the Grammar of
Graphics~\cite{wilkinson:grammar} and Polaris~\cite{stolte:polaris} (now
Tableau).

Visual encodings are defined by composing graphical primitives called
\emph{marks}~\cite{bostock:protovis}, which include \emph{arcs}, \emph{areas},
\emph{bars}, \emph{lines}, plotting \emph{symbols} and \emph{text}. Marks are
associated with datasets, and their specifications map tuple values to visual
properties such as position and color. Scales and guides (i.e., axes and
legends) are provided as first-class primitives for mapping a domain of data
values to a range of visual properties. Special \emph{group} marks serve as
containers to express nested or small multiple displays. Child marks and scales
can inherit a group's data, or draw from independent datasets.

Although interaction is a crucial component of effective
visualization~\cite{liu:mentalmodels, pike:interactionscience}, existing
declarative visualization models, including widely used tools such as
D3~\cite{bostock:d3} and ggplot2~\cite{wickham:ggplot2}, do not offer composable
primitives for interaction design. Instead, if they support interaction, they do
so through either a palette of standard techniques~\cite{bostock:protovis,
bostock:d3} or \emph{imperative} event handling callbacks. While the former
restricts expressivity, the later undoes many of the benefits of declarative
design. In particular, users are forced to contend with interaction execution
details, such as interleaved events and coordinating external state, which can
be complex and error\--prone~\cite{cooper:embedding, edwards:coherent,
myers:callbacks}.

In response, Reactive Vega introduces a model for \emph{declarative} interaction
design.

\input{vega-lang/primitives}
\input{vega-lang/examples}
\input{vega-lang/discussion}
